---
title: "VARComplexExample"
author: "Jackson Cates"
date: "9/28/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tsibble)
library(ggplot2)
library(feasts)
library(gridExtra)
library(MTS)
library(tseries)
library(forecast)
library(seastests)
```

## Data Generation

$z_{1, t} = z_{2, t-1}+\epsilon_{1,t}$

$z_{2, t} = -1.01z_{1, t-1} + 0.2z_{2, t-1}+\epsilon_{2, t}$

$z_t = \begin{pmatrix} 0 & 1 \\ -1.01 & 0.2 \end{pmatrix}z_{t-1} + \epsilon_t$

```{r}
set.seed(6)
skip = 20
length = 240
testLength = 180
noiseSd = 5

dataLength = skip + length + testLength

# Make some noise!
noise1 = rnorm(dataLength, 0, noiseSd)
noise2 = rnorm(dataLength, 0, noiseSd)

# Sets the first data point
ts1 = vector("numeric", length)
ts2 = vector("numeric", length)
ts1[1] = noise1[1]
ts2[1] = noise2[1]

# Loops though, makes linear data
for(t in 2:dataLength) {
  ts1[t] = ts2[t-1] + noise1[t] + noise1[t]
  ts2[t] = -1.01*ts1[t-1] + 0.2*ts2[t-1] + noise2[t]
}

# Takes out the testing data
test1 = ts1[(length + skip + 1):(dataLength)]
ts1 = ts1[(skip + 1):(length+skip)]
test2 = ts2[(length + skip + 1):(dataLength)]
ts2 = ts2[(skip + 1):(length+skip)]

# Turns them into a time series object
ts = as_tibble(ts1)
ts = rename(ts, "ts1" = "value")
ts[,2] = ts2
ts = rename(ts, "ts2" = "...2")
ts[,3] = 1:length
ts = rename(ts, "index" = "...3")

tsTest = as_tibble(test1)
tsTest = rename(tsTest, "ts1" = "value")
tsTest[,2] = test2
tsTest = rename(tsTest, "ts2" = "...2")
tsTest[,3] = (length + 1):(length + testLength)
tsTest = rename(tsTest, "index" = "...3")
tsTest = tsTest %>% as_tsibble(index = "index")

ts = ts %>% as_tsibble(index = "index")
```

```{r}
plot1 = ts %>% autoplot(ts1) + xlab("ts1")
plot2 = ts %>% autoplot(ts2) + xlab("ts2")
grid.arrange(plot1, plot2, nrow=2)

df1 = data.frame(time = ts$index, M = ts$ts1, isin = "ts1")
df2 = data.frame(time = ts$index, M = ts$ts2, isin = "ts2")
df = rbind(df1, df2)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("ts1 and ts2")
```

## Difference

```{r}
tsSeasonalPlot = ts %>% as.ts(start = c(1969, 2), frequency = 60)
ggseasonplot(tsSeasonalPlot[,1], polar = F) + ggtitle("Seasonal Plot: ts1")
ggseasonplot(tsSeasonalPlot[,2], polar = F) + ggtitle("Seasonal Plot: ts2")
```

### Taking the seasonal difference

$y_t^\prime = y_t - y_{t - 60}$

```{r}
tsSeasonal = ts %>% diffM(d = 60)
tsSeasonal = tsSeasonal %>% as_tibble()
tsSeasonal[,3] = 1:(length - 60)
tsSeasonal = tsSeasonal %>% as_tsibble(index = "index")

df1 = data.frame(time = tsSeasonal$index, M = tsSeasonal$ts1, isin = "ts1")
df2 = data.frame(time = tsSeasonal$index, M = tsSeasonal$ts2, isin = "ts2")
df = rbind(df1, df2)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("ts1 and ts2")

tsSeasonalPlot = tsSeasonal %>% as.ts(start = c(1969, 2), frequency = 60)
ggseasonplot(tsSeasonalPlot[,1], polar = T ) + ggtitle("Seasonal Plot: ts1 with seasonal difference")
ggseasonplot(tsSeasonalPlot[,2], polar = T ) + ggtitle("Seasonal Plot: ts2 with seasonal difference")

tsSeasonal = tsSeasonal %>% diffM(d = 60)
tsSeasonal = tsSeasonal %>% as_tibble()
tsSeasonal[,3] = 1:(length - 120)
tsSeasonal = tsSeasonal %>% as_tsibble(index = "index")

df1 = data.frame(time = tsSeasonal$index, M = tsSeasonal$ts1, isin = "ts1")
df2 = data.frame(time = tsSeasonal$index, M = tsSeasonal$ts2, isin = "ts2")
df = rbind(df1, df2)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("ts1 and ts2")

tsSeasonalPlot = tsSeasonal %>% as.ts(start = c(1969, 2), frequency = 60)
ggseasonplot(tsSeasonalPlot[,1], polar = T ) + ggtitle("Seasonal Plot: ts1 with 2nd seasonal difference")
ggseasonplot(tsSeasonalPlot[,2], polar = T ) + ggtitle("Seasonal Plot: ts2 with 2nd seasonal difference")
```


### Webel-Ollech Test

```{r}
kpss.test(tsSeasonal$ts1)
kpss.test(tsSeasonal$ts2)
```

## Fitting the model

```{r}
VARorder(tsSeasonal[,-3])
```

$z_t = \begin{pmatrix} 0 & 0.90 \\ -0.95 & 0.22 \end{pmatrix}z_{t-1} + a_t$

```{r}
m1 = VAR(tsSeasonal[,-3], p = 1)
m1R = refVAR(m1)
```


## Model Checking

### Multivariate Portmanteau Statistics

Let $R_\ell$ be the theoretical lag $\ell$ cross-correlation matrix of innovation $a_t$

$H_0$: $R_1 = \dots = R_m = 0$

$H_A$: $R_j \neq 0$ for some $1 \leq j \leq m$

```{r}
mq(m1R$residuals, lag = 12)
```

### Impulse

```{r}
VARirf(m1R$Phi, m1$Sigma)
```

## Model Forecasting

```{r}
tsPredSeasonal = VARpred(m1R, h = testLength - 120)

# Calculates the confidence interval
upperConf = tsPredSeasonal$pred + 1.96 * tsPredSeasonal$se.err
lowerConf = tsPredSeasonal$pred - 1.96 * tsPredSeasonal$se.err
```

```{r}
tsTestSeasonal = tsTest %>% diffM(d = 60) %>% diffM(d = 60)
tsTestSeasonal = tsTestSeasonal %>% as_tibble()
tsTestSeasonal[,3] = 1:(testLength - 120)
tsTestSeasonal = tsTestSeasonal %>% as_tsibble(index = "index")

df1 = data.frame(time = tsTestSeasonal$index, M = tsTestSeasonal$ts1, isin = "Actual")
df2 = data.frame(time = tsTestSeasonal$index, M = tsPredSeasonal$pred[,1], isin = "Predicted")
df3 = data.frame(time = tsTestSeasonal$index, M = upperConf[,1], isin = "Upper Bound")
df4 = data.frame(time = tsTestSeasonal$index, M = lowerConf[,1], isin = "Lower Bound")
df = rbind(df1, df2)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicted Values for ts1 without Bounds")

df = rbind(df1, df2, df3, df4)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicted Values for ts1 with Bounds")

df1 = data.frame(time = tsTestSeasonal$index, M = tsTestSeasonal$ts2, isin = "Actual")
df2 = data.frame(time = tsTestSeasonal$index, M = tsPredSeasonal$pred[,2], isin = "Predicted")
df3 = data.frame(time = tsTestSeasonal$index, M = upperConf[,2], isin = "Upper Bound")
df4 = data.frame(time = tsTestSeasonal$index, M = lowerConf[,2], isin = "Lower Bound")
df = rbind(df1, df2)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicted Values for ts2 without Bounds")

df = rbind(df1, df2, df3, df4)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicted Values for ts2 with Bounds")
```

## Inverting the Difference

$y_t = \sum_{i=0}^{t/60}y_{t-60i}^\prime + y_0$

```{r}

# # Inverts
# tsPred = vector("numeric", testLength)
# for (t in 1:testLength) {
#   
#   # Gets the current value
#   differenceSum = 0
#   for (i in 1:((t / 60) + 1)) {
#     
#     differenceSum = differenceSum + tsPredSeasonal$pred[1 + 60*(i-1),1]
#     
#   }
#   differenceSum = differenceSum + ts$ts1[length]
#   
#   print(differenceSum)
#   
# }

```


```{r}
# drawStart = 1
# 
# df1 = data.frame(time = seq(drawStart, length, length=length), M = ts$ts1, isin = "Train")
# df2 = data.frame(time = seq(length, length + testLength, length=testLength), M = tsTest$ts1, isin = "Test")
# df3 = data.frame(time = seq(length, length + testLength, length=testLength), M = pred$pred[,1], isin = "Predicted")
# df4 = data.frame(time = (length+1):(length+testLength), M = upperConf[,1], isin = "Upper Bound")
# df5 = data.frame(time = (length+1):(length + testLength), M = lowerConf[,1], isin = "Lower Bound")
# df = rbind(df1, df3)
# ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Prediction for ts1  Bounds and Test Data")
# df = rbind(df1, df2, df3)
# ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("without for ts1 Prediction Bounds")
# df = rbind(df1, df2, df3, df4, df5)
# ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("without for ts1 with Bounds")
# 
# df1 = data.frame(time = seq(drawStart, length, length=length), M = ts$ts2, isin = "Train")
# df2 = data.frame(time = seq(length, length + testLength, length=testLength), M = tsTest$ts2, isin = "Test")
# df3 = data.frame(time = seq(length, length + testLength, length=testLength), M = pred$pred[,2], isin = "Prediction")
# df4 = data.frame(time = (length+1):(length+testLength), M = upperConf[,2], isin = "Upper Bound")
# df5 = data.frame(time = (length+1):(length + testLength), M = lowerConf[,2], isin = "Lower Bound")
# df = rbind(df1, df3)
# ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicted for ts1 without Bounds and Test Data")
# df = rbind(df1, df2, df3)
# ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Prediction for ts1 without Bounds")
# df = rbind(df1, df2, df3, df4, df5)
# ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Prediction for ts1 with Bounds")
```

### MAE

```{r}
# mean(abs(tsTest$ts1 - pred$pred[,1]))
# mean(abs(tsTest$ts2 - pred$pred[,2]))
```

### MSE

```{r}
# mean((tsTest$ts1 - pred$pred[,1])*(tsTest$ts1 - pred$pred[,1]))
# mean((tsTest$ts2 - pred$pred[,2])*(tsTest$ts2 - pred$pred[,2]))
```
