---
title: "VarLinearExample"
author: "Jackson Cates"
date: "9/26/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tsibble)
library(ggplot2)
library(feasts)
library(gridExtra)
library(MTS)
library(tseries)
```

## Data Generation

$z_{1, t} = 0.8z_{1, t-1} + 0.6z_{2, t-1}+0.1t+ \epsilon_{1,t}+0.12\epsilon_{1,t-1}$

$z_{2, t} = 0.2z_{1, t-1} + 0.3z_{2, t-1}+\epsilon_{2, t}+0.1\epsilon_{1,t-1} + 0.025\epsilon_{2, t-1}$

$z_t = \begin{pmatrix} 0.1 \\ 0 \end{pmatrix}t + \begin{pmatrix} 0.8 & 0.6 \\ 0.2 & 0.3 \end{pmatrix}z_{t-1} + \epsilon_t + \begin{pmatrix} 0.12 & 0 \\ 0.1 & 0.025 \end{pmatrix}\epsilon_{t-1}$

```{r}
set.seed(6)
skip = 10
length = 100
testLength = 20
noiseSd = 5

dataLength = skip + length + testLength

# Make some noise!
noise1 = rnorm(dataLength, 0, noiseSd)
noise2 = rnorm(dataLength, 0, noiseSd)

# Sets the first data point
ts1 = vector("numeric", length)
ts2 = vector("numeric", length)
ts1[1] = noise1[1]
ts2[1] = noise2[1]

# Loops though, makes linear data
for(t in 2:dataLength) {
  ts1[t] = 0.8*ts1[t-1] + 0.6*ts2[t-1] + 0.1*t + noise1[t] + 0.12*noise1[t-1]
  ts2[t] = 0.2*ts1[t-1] + 0.3*ts2[t-1] + noise2[t] + 0.1*noise1[t-1] + 0.025*noise2[t-1]
}

# Takes out the testing data
test1 = ts1[(length + skip + 1):(dataLength)]
ts1 = ts1[(skip + 1):(length+skip)]
test2 = ts2[(length + skip + 1):(dataLength)]
ts2 = ts2[(skip + 1):(length+skip)]

# Turns them into a time series object
ts = as_tibble(ts1)
ts = rename(ts, "ts1" = "value")
ts[,2] = ts2
ts = rename(ts, "ts2" = "...2")
ts[,3] = 1:length
ts = rename(ts, "index" = "...3")

tsTest = as_tibble(test1)
tsTest = rename(tsTest, "ts1" = "value")
tsTest[,2] = test2
tsTest = rename(tsTest, "ts2" = "...2")
tsTest[,3] = (length + 1):(length + testLength)
tsTest = rename(tsTest, "index" = "...3")
tsTest = tsTest %>% as_tsibble(index = "index")

ts = ts %>% as_tsibble(index = "index")
```

```{r}
plot1 = ts %>% autoplot(ts1) + xlab("ts1")
plot2 = ts %>% autoplot(ts2) + xlab("ts2")
grid.arrange(plot1, plot2, nrow=2)

df1 = data.frame(time = ts$index, M = ts$ts1, isin = "ts1")
df2 = data.frame(time = ts$index, M = ts$ts2, isin = "ts2")
df = rbind(df1, df2)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("ts1 and ts2")
```

## Differencing

We are going to take the change between consecutive observations

$y_t^\prime = y_t - y_{t-1}$

```{r}
# Takes a lagged difference of 1
tsDiff = ts %>% diffM()
tsDiff[,3] = 1:(length-1)
tsDiff = tsDiff %>% as_tibble()
tsDiff = tsDiff %>% as_tsibble(index = "index")

df1 = data.frame(time = tsDiff$index, M = tsDiff$ts1, isin = "ts1")
df2 = data.frame(time = tsDiff$index, M = tsDiff$ts2, isin = "ts2")
df = rbind(df1, df2)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("ts1 and ts2")

kpss.test(tsDiff$ts1)
kpss.test(tsDiff$ts2)
```

```{r}
VARorder(tsDiff[,-3])
```

## Fitting the model

$z_t = \begin{pmatrix} 2.06 \\ 0 \end{pmatrix} + \begin{pmatrix} 0 & 0.32 \\ 0.35 & -0.42 \end{pmatrix}z_{t-1} + \begin{pmatrix} -0.10 & 0 \\ 0 & -0.29 \end{pmatrix}z_{t-2} +  a_t$

```{r}
# Does LS estimation of the model
m1 = VAR(tsDiff[,-3], 2)
m1R = refVAR(m1)
```

## Model Checking

### Multivariate Portmanteau Statistics

Let $R_\ell$ be the theoretical lag $\ell$ cross-correlation matrix of innovation $a_t$

$H_0$: $R_1 = \dots = R_m = 0$

$H_A$: $R_j \neq 0$ for some $1 \leq j \leq m$

```{r}
mq(m1R$residuals)
```

### Impulse

The first graph is the impulse response, while the second is accumulated response.

```{r}
VARirf(m1R$Phi, m1$Sigma)
```


## Model Forecasting

```{r}
pred = VARpred(m1R, h = testLength)

# Calculates the confidence interval
upperConfDiff = pred$pred + 1.96 * pred$se.err
lowerConfDiff = pred$pred - 1.96 * pred$se.err
```

```{r}
testDiff = tsTest %>% diffM()
testDiff[,3] = 1:(testLength-1)
testDiff = testDiff %>% as_tibble()
testDiff = testDiff %>% as_tsibble(index = "index")

tsPredDiff = as_tibble(pred$pred[,1])
tsPredDiff = rename(tsPredDiff, "ts1" = "value")
tsPredDiff[,2] = pred$pred[,2]
tsPredDiff = rename(tsPredDiff, "ts2" = "...2")
tsPredDiff[,3] = 1:testLength
tsPredDiff = rename(tsPredDiff, "index" = "...3")
tsPredDiff = tsPredDiff %>% as_tsibble(index = "index")

df1 = data.frame(time = testDiff$index, M = testDiff$ts1, isin = "Actual")
df2 = data.frame(time = tsPredDiff$index, M = tsPredDiff$ts1, isin = "Predicited")
df3 = data.frame(time = 1:testLength, M = upperConfDiff[,1], isin = "Upper Bound")
df4 = data.frame(time = 1:testLength, M = lowerConfDiff[,1], isin = "Lower Bound")
df = rbind(df1, df2, df3, df4)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicited of Differenced Values for ts1")

df1 = data.frame(time = testDiff$index, M = testDiff$ts2, isin = "Actual")
df2 = data.frame(time = tsPredDiff$index, M = tsPredDiff$ts2, isin = "Predicited")
df3 = data.frame(time = 1:testLength, M = upperConfDiff[,2], isin = "Upper Bound")
df4 = data.frame(time = 1:testLength, M = lowerConfDiff[,2], isin = "Lower Bound")
df = rbind(df1, df2, df3, df4)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicited of Differenced Values for ts2")
```

## Inverting the Differencing

$y_t = \sum_{i=1}^ty_i\prime + y_0$

```{r}
# Accumlates all of the transfomred values
tsPred = tsPredDiff[,-3] %>% cumsum() %>% as_tibble()
upperConf = matrix(nrow = testLength, ncol = 2)
upperConf[,1] = upperConfDiff[,1] %>% cumsum()
upperConf[,2] = upperConfDiff[,2] %>% cumsum()
lowerConf = matrix(nrow = testLength, ncol = 2)
lowerConf[,1] = lowerConfDiff[,1] %>% cumsum()
lowerConf[,2] = lowerConfDiff[,2] %>% cumsum()

# Adds the intercepts
tsPred$ts1 = tsPred$ts1 + ts$ts1[length]
tsPred$ts2 = tsPred$ts2 + ts$ts2[length]
tsPred[,3] = (length + 1):(length + testLength)
tsPred = rename(tsPred, "index" = "...3")
tsPred = tsPred %>% as_tsibble(index = "index")
upperConf[,1] = upperConf[,1] + ts$ts1[length]
upperConf[,2] = upperConf[,2] + ts$ts2[length]
lowerConf[,1] = lowerConf[,1] + ts$ts1[length]
lowerConf[,2] = lowerConf[,2] + ts$ts2[length]

df1 = data.frame(time = tsTest$index, M = tsTest$ts1, isin = "Actual")
df2 = data.frame(time = tsPred$index, M = tsPred$ts1, isin = "Predicited")
df3 = data.frame(time = (length+1):(length+testLength), M = upperConf[,1], isin = "Upper Bound")
df4 = data.frame(time = (length+1):(length + testLength), M = lowerConf[,1], isin = "Lower Bound")
df = rbind(df1, df2, df3, df4)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicited of Non-Differenced Values for ts1 with Bounds")
df = rbind(df1, df2)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicited of Non-Differenced Values for ts1 without Bounds")

df1 = data.frame(time = tsTest$index, M = tsTest$ts2, isin = "Actual")
df2 = data.frame(time = tsPred$index, M = tsPred$ts2, isin = "Predicited")
df3 = data.frame(time = (length+1):(length+testLength), M = upperConf[,2], isin = "Upper Bound")
df4 = data.frame(time = (length+1):(length + testLength), M = lowerConf[,2], isin = "Lower Bound")
df = rbind(df1, df2, df3, df4)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicited of Non-Differenced Values for ts2 with Bounds")
df = rbind(df1, df2)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicited of Non-Differenced Values for ts2 without Bounds")
```

```{r}
drawStart = 1

df1 = data.frame(time = seq(drawStart, length, length=length), M = ts$ts1, isin = "Train")
df2 = data.frame(time = seq(length, length + testLength, length=testLength), M = tsTest$ts1, isin = "Test")
df3 = data.frame(time = seq(length, length + testLength, length=testLength), M = tsPred$ts1, isin = "Predicited")
df4 = data.frame(time = (length+1):(length+testLength), M = upperConf[,1], isin = "Upper Bound")
df5 = data.frame(time = (length+1):(length + testLength), M = lowerConf[,1], isin = "Lower Bound")
df = rbind(df1, df2, df3, df4, df5)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicition for ts1 with Bounds")
df = rbind(df1, df2, df3)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicition for ts1 witout Bounds")

df1 = data.frame(time = seq(drawStart, length, length=length), M = ts$ts2, isin = "Train")
df2 = data.frame(time = seq(length, length + testLength, length=testLength), M = tsTest$ts2, isin = "Test")
df3 = data.frame(time = seq(length, length + testLength, length=testLength), M = tsPred$ts2, isin = "Predicited")
df4 = data.frame(time = (length+1):(length+testLength), M = upperConf[,2], isin = "Upper Bound")
df5 = data.frame(time = (length+1):(length + testLength), M = lowerConf[,2], isin = "Lower Bound")
df = rbind(df1, df2, df3, df4, df5)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicition for ts2")
df = rbind(df1, df2, df3)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predicition for ts2 witout Bounds")
```


### MAE

```{r}
mean(abs(tsTest$ts1 - tsPred$ts1))
mean(abs(tsTest$ts2 - tsPred$ts2))
```

### MSE

```{r}
mean((tsTest$ts1 - tsPred$ts1)*(tsTest$ts1 - tsPred$ts1))
mean((tsTest$ts2 - tsPred$ts2)*(tsTest$ts2 - tsPred$ts2))
```
