---
title: "VAR Stationary Example"
author: "Jackson Cates"
date: "9/15/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tsibble)
library(ggplot2)
library(feasts)
library(gridExtra)
library(MTS)
```


## Data Generation

The model I am going to simulate is as follows:

$ts1_t = 0.3*ts1_{t-1} + \epsilon_1$

$ts2_t = 0.4*ts2_{t-1} + ts1_{t-1} + \epsilon_2$

Which results in the following model

$\begin{pmatrix}ts1_t\\ts2_t\end{pmatrix}=\begin{pmatrix}0.3 &0\\1&0.4\end{pmatrix}\begin{pmatrix} ts1_{t-1} \\ts2_{t-1}\end{pmatrix}+\begin{pmatrix} \epsilon_1\\\epsilon_2\end{pmatrix}$

```{r}
# Generates the two vectors
set.seed(14)
skip = 200 #10
length = 520 #50
testLength = 20
noiseSd = 1

# Simluates the model
noise = rnorm(length, 0, noiseSd)
ts1 = vector("numeric", length)
ts1[1] = noise[1] + 1
for(t in 2:length) {
  ts1[t] = 0.3*ts1[t - 1] + noise[t]
}

noise = rnorm(length, 0, noiseSd)
ts2 = vector("numeric", length)
ts2[1] = noise[1] + 1
for(t in 2:length) {
  ts2[t] = 0.4*ts2[t - 1] + 1*ts1[t - 1] + noise[t]
}

# noise = rnorm(length, 0, noiseSd)
# ts1 = ts1 + noise
# noise = rnorm(length, 0, noiseSd)
# ts2 = ts2 + noise

# pi = matrix(c(0.3, 1.0, 0, 0.4), 2, 2)
# sig = matrix(c(4,0.8,0.8,1),2,2)
# genData = VARMAsim(length, arlags = c(1), malags = NULL, cnst = NULL, phi = pi, theta = NULL, sigma = sig)
# 
# ts1 = genData$series[,1]
# ts2 = genData$series[,2]

# Takes out the testing data
test1 = ts1[(length-testLength):(length-1)]
ts1 = ts1[(skip+1):(length-testLength)]
test2 = ts2[(length-testLength):(length-1)]
ts2 = ts2[(skip+1):(length-testLength)]

length = length - testLength - skip

# Turns them into a time series object
ts = as_tibble(ts1)
ts = rename(ts, "ts1" = "value")
ts[,2] = ts2
ts = rename(ts, "ts2" = "...2")
ts[,3] = 1:length
ts = rename(ts, "index" = "...3")

ts = ts %>% as_tsibble(index = "index")
```

```{r}
plot1 = ts %>% autoplot(ts1) + xlab("ts1")
plot2 = ts %>% autoplot(ts2) + xlab("ts2")
grid.arrange(plot1, plot2, nrow=2)

df1 = data.frame(time = ts$index, M = ts$ts1, isin = "ts1")
df2 = data.frame(time = ts$index, M = ts$ts2, isin = "ts2")
df = rbind(df1, df2)
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("ts1 and ts2")
```

\newpage

## Model Selection

We are going to try to select a $VAR(p)$ model. The model is:

$z_t = \phi_0 + \sum^p_{i=1}\phi_iz_{t-i} + a_t$

- $k$ is the number of time series we have
- $\phi_0$ is a $k$ dimensional constant vector
- $\phi_i$ is a $k$ by $k$ matrix
- $a_t$ is a sequence of independent and identically distributed random vectors with mean zero and covariance matrix $\Sigma_a$

### Order Selection with the Sequential likelihood ratio test:

What we are going to do for our order selection is compare $VAR(\ell)$ with $VAR(\ell - 1)$

$H_0$: $\phi_\ell=0$

$H_A$: $\phi_\ell \neq 0$

```{r}
# Does the order test 
VARorder(ts[,-3])
```

As seen above, we should proceed with a $VAR(1)$ model

\newpage

## Fitting the model

```{r}
# Does LS estimation of the model
m1 = VAR(ts[,-3], 1)
```

From the output above, we get the following model:

$z_t=\begin{pmatrix}-0.04\\-0.05\end{pmatrix}+\begin{pmatrix}0.30&0.03\\0.929&0.44\end{pmatrix}z_{t-1}+a_t$

Some things to note:

- Granger Causality is low from ts2 to ts1 (with value of 0.04)
- Coefficients are really similar to our simulation

\newpage

## Model Checking

### Stationarity

Its turns out that to test if a series is stationary, we can solve the following determinate:

$|I_k - \Phi_1B| = 0$

and if the absolute value of the solutions are greater than 1, it is stationary!

For $VAR(1)$ models, the solutions of $B$ is simply the reciprocal of the eigenvalues of $\Phi_1$

```{r}
eigen(m1$Phi)[1]
```

So our series is stationary!

### Multivariate Portmanteau Statistics

Let $R_\ell$ be the theoretical lag $\ell$ cross-correlation matrix of innovation $a_t$

$H_0$: $R_1 = \dots = R_m = 0$

$H_A$: $R_j \neq 0$ for some $1 \leq j \leq m$

```{r}
mq(m1$residuals)
```

### Impulse

The first graph is the impulse response, while the second is accumulated response.

```{r}
VARirf(m1$Phi, m1$Sigma)
```

\newpage

## Model Forecasting

### Static Forecasting

```{r}
pred = VARpred(m1, h = testLength)

# Calculates the confidence interval
upperConf = pred$pred + 1.96 * pred$se.err
lowerConf = pred$pred - 1.96 * pred$se.err
```

```{r}
drawLength = testLength
drawStart = length - drawLength

## wrap data into a data.frame
df1 = data.frame(time = seq(drawStart,length,length=drawLength), M = ts$ts1[(drawStart+1):length], isin = "observations")
df2 = data.frame(time = seq(length,length+testLength,length=testLength), M = pred$pred[1] , isin = "forecast")
df3 = data.frame(time = seq(length,length+testLength,length=testLength), M = upperConf[1],isin = "upper_bound")
df4 = data.frame(time = seq(length,length+testLength,length=testLength), M = lowerConf[1], isin = "lower_bound")
df5 = data.frame(time = seq(length,length+testLength,length=testLength), M = test1 , isin = "actual")
df = rbind(df1, df2, df3, df4, df5)

## ggplot object 
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predictions for ts1") + geom_smooth(aes(x=time, y=M, ymax=upperConf[1], ymin=lowerConf[1]), colour="yellow", data=df2, stat='identity')

## wrap data into a data.frame
df1 = data.frame(time = seq(drawStart,length,length=drawLength), M = ts$ts2[(drawStart+1):length], isin = "observations")
df2 = data.frame(time = seq(length,length+testLength,length=testLength), M = pred$pred[2] , isin = "forecast")
df3 = data.frame(time = seq(length,length+testLength,length=testLength), M = upperConf[2],isin = "upper_bound")
df4 = data.frame(time = seq(length,length+testLength,length=testLength), M = lowerConf[2], isin = "lower_bound")
df5 = data.frame(time = seq(length,length+testLength,length=testLength), M = test2, isin = "actual")
df = rbind(df1, df2, df3, df4, df5)

## ggplot object 
ggplot(df, aes(x = time, y = M, color = isin)) + geom_line() + ggtitle("Predictions for ts2") + geom_smooth(aes(x=time, y=M, ymax=upperConf[2], ymin=lowerConf[2]), colour="yellow", data=df2, stat='identity')
```

### MAE

```{r}
mean(abs(test1 - pred$pred[1]))
mean(abs(test2 - pred$pred[2]))
```

### MSE

```{r}
mean((test1 - pred$pred[1])*(test1 - pred$pred[1]))
mean((test2 - pred$pred[2])*(test2 - pred$pred[2]))
```