---
title: "Multivariate Modeling Example"
author: "Jackson Cates"
date: "9/5/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r, warning=FALSE, message=FALSE}
library(readr)
library(marima)
library(tsibble)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tidyverse)
library(fable)
library(feasts)
library(urca)
library(gridExtra)
library(forecast)
library(tseries)
library(MTS)
library(vars)
```

## Data Set Summary
```{r}
head(EuStockMarkets)
tail(EuStockMarkets)

# Seperates the data into training and testing data
train = subset(EuStockMarkets, start = 1, end = nrow(EuStockMarkets) - 14)
test = subset(EuStockMarkets, start = nrow(EuStockMarkets) - 14, end = nrow(EuStockMarkets))

# Makes the time plot
train %>% autoplot() + ylab("Stock Value")
```

\pagebreak

## Testing Stationary
```{r}
plot1 = ggseasonplot(train[,"DAX"], year.labels = T, main = "Seasonal Plot: DAX", ylab = "Stock Value")
plot2 = ggseasonplot(train[,"SMI"], year.labels = T, main = "Seasonal Plot: SMI", ylab = "Stock Value")
plot3 = ggseasonplot(train[,"CAC"], year.labels = T, main = "Seasonal Plot: CAC", ylab = "Stock Value")
plot4 = ggseasonplot(train[,"FTSE"], year.labels = T, main = "Seasonal Plot: FTSE", ylab = "Stock Value")
grid.arrange(plot1, plot2, plot3, plot4, ncol = 2, nrow = 2)
```

### Unitroot Test
$H_O:$ The data is stationary

$H_A:$ The data is not stationary

```{r}
kpss.test(train[,"DAX"])[3]
kpss.test(train[,"SMI"])[3]
kpss.test(train[,"CAC"])[3]
kpss.test(train[,"FTSE"])[3]
```

\pagebreak

## Differencing

We are going to take the change between consecutive observations

$y_t^\prime = y_t - y_{t-1}$

```{r}
trainDiff = train %>% diffM()

plot.ts(trainDiff)
autoplot(ts(trainDiff))

trainDiff = trainDiff %>% diffM()

plot.ts(trainDiff)
autoplot(ts(trainDiff))
```

### Unitroot Test
$H_O:$ The data is stationary

$H_A:$ The data is not stationary

```{r}
kpss.test(trainDiff[,"DAX"])[3]
kpss.test(trainDiff[,"SMI"])[3]
kpss.test(trainDiff[,"CAC"])[3]
kpss.test(trainDiff[,"FTSE"])[3]
```

\pagebreak

## Selecting p and q

***Eccm*** function in R gives a matrix of multivariate Ljung-Box statistics of a vector time series.

$H_O:$ The model does not show a lack of fit (that the autocorrelations (for the chosen lags) in the population from which the sample is taken are all zero).

$H_A:$ The model does show a lack of fit

```{r}
Eccm(trainDiff)
```

\pagebreak

## Fitting the model

| p | q |   aic  |   bic   | refAic | refBic | ref2Aic | ref2Bic |
|---|---|--------|---------|--------|--------|---------|---------|
| 2 | 2 |25.23482|25.43837 |25.54427|25.71489|NA       |NA       |
| 3 | 1 |25.31184|25.51539 |25.32326|25.41007|25.61058 |25.79916 |
| 3 | 2 |58.92585|59.17729 |41.12731|41.36678|25.65966 |25.80334 |


```{r}
fit = VARMA(trainDiff, p = 2, q = 2)
```

\pagebreak

## Model Checking

```{r}
MTSdiag(fit)
phi=fit$Phi; theta=fit$Theta; sig=fit$Sigma 
VARMAirf(Phi=phi,Theta=theta,Sigma=sig,orth=F)
```

